---
title: "MLB Statistical Analysis"
author: "Santiago Alvarez Barbosa, Santiago Diaz Jinete, Nicolás Velandia Sanabria"
date: "11/20/2021"
output: html_document
---
Loading Packages...
```{r}
library(ggplot2)
library(ggthemes)
library(MASS)
library(caret)
library(klaR)
```

Loading 2015-2019 Data and Data Cleaning:
Reads three CSV files for batting, errors and pitches for seasons between 2015 and 2019. 
Specific columns from the batting and pitching data frames are chosen and renamed.
Combines these modified data frames with a specific column from the error data frame, resulting in `data_2015` data frame.
Two rows removed and new column added that holds the value of the season.

```{r}
bat_2015 <- read.csv("Batting_2015.csv")
bat_2015 <- bat_2015[,c(1,4,6,7,9,10,11,12,14,17)]
colnames(bat_2015)[colnames(bat_2015) == "SO"] <- "SOA"
error_2015 <- read.csv("Error_2015.csv")
E <- error_2015[,12]
pitch_2015 <- read.csv("Pitching_2015.csv")
pitch_2015 <- pitch_2015[,c(4,8,13,17,20,21,23,24)]
colnames(pitch_2015)[colnames(pitch_2015) %in% c("H","HR")] <- c("HA","HRA")
data_2015 = cbind(bat_2015,pitch_2015,E)
data_2015 = data_2015[c(-31,-32),]
data_2015$Ssn <- 2015 
```

```{r}
bat_2016 <- read.csv("Batting_2016.csv")
bat_2016 <- bat_2016[,c(1,4,6,7,9,10,11,12,14,17)]
colnames(bat_2016)[colnames(bat_2016) == "SO"] <- "SOA"
error_2016 <- read.csv("Error_2016.csv")
E <- error_2016[,12]
pitch_2016 <- read.csv("Pitching_2016.csv")
pitch_2016 <- pitch_2016[,c(4,8,13,17,20,21,23,24)]
colnames(pitch_2016)[colnames(pitch_2016) %in% c("H","HR")] <- c("HA","HRA")
data_2016 = cbind(bat_2016,pitch_2016,E)
data_2016 = data_2016[c(-31,-32),]
data_2016$Ssn <- 2016 
```

```{r}
bat_2017 <- read.csv("Batting_2017.csv")
bat_2017 <- bat_2017[,c(1,4,6,7,9,10,11,12,14,17)]
colnames(bat_2017)[colnames(bat_2017) == "SO"] <- "SOA"
error_2017 <- read.csv("Error_2017.csv")
E <- error_2017[,12]
pitch_2017 <- read.csv("Pitching_2017.csv")
pitch_2017 <- pitch_2017[,c(4,8,13,17,20,21,23,24)]
colnames(pitch_2017)[colnames(pitch_2017) %in% c("H","HR")] <- c("HA","HRA")
data_2017 = cbind(bat_2017,pitch_2017,E)
data_2017 = data_2017[c(-31,-32),]
data_2017$Ssn <- 2017 
```

```{r}
bat_2018 <- read.csv("Batting_2018.csv")
bat_2018 <- bat_2018[,c(1,4,6,7,9,10,11,12,14,17)]
colnames(bat_2018)[colnames(bat_2018) == "SO"] <- "SOA"
error_2018 <- read.csv("Error_2018.csv")
E <- error_2018[,12]
pitch_2018 <- read.csv("Pitching_2018.csv")
pitch_2018 <- pitch_2018[,c(4,8,13,17,20,21,23,24)]
colnames(pitch_2018)[colnames(pitch_2018) %in% c("H","HR")] <- c("HA","HRA")
data_2018 = cbind(bat_2018,pitch_2018,E)
data_2018 = data_2018[c(-31,-32),]
data_2018$Ssn <- 2018 
```

```{r}
bat_2019 <- read.csv("Batting_2019.csv")
bat_2019 <- bat_2019[,c(1,4,6,7,9,10,11,12,14,17)]
colnames(bat_2019)[colnames(bat_2019) == "SO"] <- "SOA"
error_2019 <- read.csv("Error_2019.csv")
E <- error_2019[,12]
pitch_2019 <- read.csv("Pitching_2019.csv")
pitch_2019 <- pitch_2019[,c(4,8,13,17,20,21,23,24)]
colnames(pitch_2019)[colnames(pitch_2019) %in% c("H","HR")] <- c("HA","HRA")
data_2019 = cbind(bat_2019,pitch_2019,E)
data_2019 = data_2019[c(-31,-32),]
data_2019$Ssn <- 2019 
```
Appends a "Ps" column to the previously made data frames.
The values in this new column indicate whether a team participated in the playoffs or not.

```{r}
post_2019 = c("Non-Playoff","Playoff","Non-Playoff","Non-Playoff","Non-Playoff","Non-Playoff","Non-Playoff","Non-Playoff","Non-Playoff","Non-Playoff","Playoff","Non-Playoff","Non-Playoff","Playoff","Non-Playoff","Playoff","Playoff","Non-Playoff","Playoff","Playoff","Non-Playoff","Non-Playoff","Non-Playoff","Non-Playoff","Non-Playoff","Playoff","Playoff","Non-Playoff","Non-Playoff","Playoff")
post_2018 = c("Non-Playoff","Playoff","Non-Playoff","Playoff","Playoff","Non-Playoff","Non-Playoff","Playoff","Playoff","Non-Playoff","Playoff","Non-Playoff","Non-Playoff","Playoff","Non-Playoff","Playoff","Non-Playoff","Non-Playoff","Playoff","Playoff","Non-Playoff","Non-Playoff","Non-Playoff","Non-Playoff","Non-Playoff","Non-Playoff","Non-Playoff","Non-Playoff","Non-Playoff","Non-Playoff")
post_2017 = c("Playoff","Non-Playoff","Non-Playoff","Playoff","Playoff","Non-Playoff","Non-Playoff","Playoff","Playoff","Non-Playoff","Playoff","Non-Playoff","Non-Playoff","Playoff","Non-Playoff","Non-Playoff","Playoff","Non-Playoff","Playoff","Non-Playoff","Non-Playoff","Non-Playoff","Non-Playoff","Non-Playoff","Non-Playoff","Non-Playoff","Non-Playoff","Non-Playoff","Non-Playoff","Playoff")
post_2016 = c("Non-Playoff","Non-Playoff","Playoff","Playoff","Playoff","Non-Playoff","Non-Playoff","Playoff","Non-Playoff","Non-Playoff","Non-Playoff","Non-Playoff","Non-Playoff","Playoff","Non-Playoff","Non-Playoff","Non-Playoff","Playoff","Non-Playoff","Non-Playoff","Non-Playoff","Non-Playoff","Non-Playoff","Non-Playoff","Playoff","Non-Playoff","Non-Playoff","Playoff","Playoff","Playoff")
post_2015 = c("Non-Playoff","Non-Playoff","Non-Playoff","Non-Playoff","Playoff","Non-Playoff","Non-Playoff","Non-Playoff","Non-Playoff","Non-Playoff","Playoff","Playoff","Non-Playoff","Playoff","Non-Playoff","Non-Playoff","Non-Playoff","Playoff","Playoff","Non-Playoff","Non-Playoff","Playoff","Non-Playoff","Non-Playoff","Non-Playoff","Playoff","Non-Playoff","Playoff","Playoff","Non-Playoff")
data_2019 = cbind(data_2019,post_2019)
colnames(data_2019)[colnames(data_2019) == "post_2019"] <- "Ps"
data_2018 = cbind(data_2018,post_2018)
colnames(data_2018)[colnames(data_2018) == "post_2018"] <- "Ps"
data_2017 = cbind(data_2017,post_2017)
colnames(data_2017)[colnames(data_2017) == "post_2017"] <- "Ps"
data_2016 = cbind(data_2016,post_2016)
colnames(data_2016)[colnames(data_2016) == "post_2016"] <- "Ps"
data_2015 = cbind(data_2015,post_2015)
colnames(data_2015)[colnames(data_2015) == "post_2015"] <- "Ps"
```

```{r}
data = rbind(data_2019,data_2018,data_2017,data_2016,data_2015)
head(data)
```

#### Análisis de Componentes Principales 

```{r}
summary(data)
```
Let's remember that PCA only accepts continuous variables, meaning that variables like *Tm* and *Ssn*, which represent the team and the season respectively, will not be taken into account.

Applying PCA...

```{r}
data.pca <- prcomp(data[,c(-1,-20,-21)], scale = TRUE)
```

Having the Principal Components (PCs), let's examine the cumulative proportion and the eigenvalues to determine the number of components we should choose.

```{r}
summary(data.pca)
```
So, we determine that an appropriate number of principal components is 7. This is because the majority of the eigenvalues are equal to or greater than one, and additionally, 84.4% of the total variance is accumulated in the first 7 PCs.

We can verify this conclusion with a Scree Plot.

```{r}
variance = data.pca$sdev^2
qplot(c(1:18), variance) +
  geom_line(color='darkred') +
  geom_point(size=2,color='darkblue')+
  xlab("Principal Component") +
  ylab("Variance") +
  ggtitle("Scree Plot for MLB Data") +
  ylim(0, 6) +
  theme_igray() 
```
As you can see, in the 8th principal component, the variance begins to level off, so choosing 7 principal components is a suitable decision.

Next, the selected 7 principal components are presented:

```{r}
data.pca$rotation[,1:7]
```

#### Análisis Discriminante 

To perform a discriminant analysis, we will use the previously derived principal components since we already know they are uncorrelated.

We will utilize the data from the 2015-2018 seasons as training data for the model. Consequently, we will observe how the model classifies the teams in the 2019 season.

```{r}
new.data <- cbind(data[c("Ps","Ssn")],data.pca$x[,1:7])
training = (new.data$Ssn < 2019)
testing = !training

training_data = new.data[training,]
testing_data = new.data[testing,]

post_season = new.data$Ps[testing]
```
Now, because $\Sigma_1 \neq \Sigma_2$ for our Playoff and Non-Playoff populations, we will use a quadratic discriminant analysis.

First, we fit the model:

```{r}
qda_model <- qda(Ps~.,data=training_data,type = "C-Classification")
```
We validate our model using the testing dataset:

```{r}
qda_pred <- predict(qda_model, testing_data)
qda_pred_post_season = qda_pred$class
```
We obtain the Confusion Matrix:

```{r}
cmatrix_lda<-table(qda_pred_post_season,post_season, dnn = c("Actual Group", "Predicted Group"))
confusionMatrix(cmatrix_lda)
```
Finally, we calculate the error of misclassification.

```{r}
mean(qda_pred_post_season!= post_season)
```
So, as can be seen, the proposed model is 83.3% accurate in classifying teams into the *Playoffs* and *Non-Playoffs* populations. Additionally, we found that the error of misclassifying one of these teams is 16.6%, which is relatively small.

Therefore, we have the following table that compares the results obtained from the Discriminant Analysis with the actual values:
```{r}
data_2019_pred = cbind(data_2019[c("Tm","Ssn","Ps")],qda_pred_post_season)
colnames(data_2019_pred)[colnames(data_2019_pred) == "Tm"] <- "Team"
colnames(data_2019_pred)[colnames(data_2019_pred) == "Ssn"] <- "Season"
colnames(data_2019_pred)[colnames(data_2019_pred) == "Ps"] <- "True Population"
colnames(data_2019_pred)[colnames(data_2019_pred) == "qda_pred_post_season"] <- "Predicted Population"
data_2019_pred
```
Finally, we present the results using the "partimat" function, which allows visualizing the classifications of 2 variables at once.

```{r}
testing_data$Ps = as.factor(testing_data$Ps)
partimat(testing_data[,3:7],testing_data$Ps, data=testing_data, method = "qda")
```

